---
title: "COSICC_DA_group and COSICC_DA_lineage"
author: "Magdalena Strauss"
date: "`r Sys.Date()`"
output: pdf_document
---

This tutorial illustrates how to use COSICC_DA_group with SingleCellExperiments and with 
SeuratObjects. 

```{r}

knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  error = FALSE
)

suppressPackageStartupMessages(library(COSICC))
```

# COSICC_DA_group for SingleCellExperiments

First, we simulate a chimera-style data set, where cells have a fluorescent marker tdTomato or not. 

```{r}
sim_data_sce <- simulate_sce(seed = 10)
```

The function above simulated a knockout and a wild-type chimera data set. 

```{r}
sce_knockout <- sim_data_sce$case
sce_WT <- sim_data_sce$control
```

Let's have a look at the simulated data sets:

```{r}
colData(sce_knockout)
```


```{r}
colData(sce_WT)
```
The tdTomato coloumn contains the values "pos" and "neg".

To use COSICC_DA_group we need to rename them to TRUE and FALSE. 

```{r}
sce_WT$tdTomato <- sce_WT$tdTomato == "pos" 
sce_knockout$tdTomato <- sce_knockout$tdTomato == "pos"
```

Furthermore, we need to rename the colData. 

```{r}
names(colData(sce_WT))[names(colData(sce_WT))== "tdTomato"] <- "marked"
names(colData(sce_WT))[names(colData(sce_WT)) == "celltype"] <- "cell_type"

names(colData(sce_knockout))[names(colData(sce_knockout))== "tdTomato"] <- "marked"
names(colData(sce_knockout))[names(colData(sce_knockout)) == "celltype"] <- "cell_type"
```

We also make sure that the cells from the WT and knockout data sets have different names. 

```{r}
colnames(sce_WT) <- paste0(colnames(sce_WT),"_WT")
colnames(sce_knockout) <- paste0(colnames(sce_knockout),"_knockout")
```

Now the data look as follows:

```{r}
head(colData(sce_WT))
```

We can now COSICC_DA_group to identify depletion and/or enrichment of cell types for the tdTomato positive cells in the knockout chimeras. 

```{r}
DA_result_sce <- COSICC_DA_group(
    sce_case=sce_knockout, 
    sce_control=sce_WT
) 
```
The figure above illustrates the cell types Group1, Group2 and Group3 are depleted for the tdTomato positive group in the knockout data set compared to the wild-type data set. 

The output of the function COSICC_DA_group is a  data frame with the following columns. 

| Column Name   | Description                                                                 |
|---------------|-----------------------------------------------------------------------------|
| `cell_type`   | Cell type name.                                                              |
| `FDR`         | FDR computed using the Benjamini-Hochberg method.                             |
| `odds_ratio`  | Odds ratio of enrichment/depletion for each group of cells or cell type.    |
| `sig`         | Significance status: "enriched", "depleted", or "not significant".           |

Below we print the output. 

```{r}
DA_result_sce
```

# COSICC_DA_group for SeuratObject

First, we simulate a chimera-style data set, where cells have a fluorescent marker tdTomato or not. 

```{r}
library(Seurat)
sim_data_seurat <- simulate_seurat(seed = 10)
seurat_knockout <- sim_data_seurat$case
seurat_WT <- sim_data_seurat$control
```

Now we create SingleCellExperiments.

```{r}
sce_WT <- SingleCellExperiment(assays=list(counts=seurat_WT@assays$RNA),colData=seurat_WT@meta.data)
sce_knockout <- SingleCellExperiment(assays=list(counts=seurat_knockout@assays$RNA),colData=seurat_knockout@meta.data)
```

Now you can use COSICC_DA_group as described in the section on COSICC_DA_group for SingleCellExperiments above. 

# COSICC_DA_lineage

To illustrate COSICC_DA_lineage, we simulate lineage scores. In application in development, these scores might be scores indicating probabilities of a cells turning into each lineage. We use example data from the package. 

```{r}
data(package="COSICC")
```

This shows that the package contains the following example data sets:

lineage_scores           
sce_DA_lineage_case      
sce_DA_lineage_control  

Note that sce_DA_lineage_case and sce_DA_lineage_control have a slot cell in the 
colData. This is the cell ID, or the ID of the mapped cell in a reference atlas. It
needs to be identical to the id column of the lineage_scores data frame (see below).

```{r}
head(sce_DA_lineage_case)
```

The scores look as follows:

```{r}
head(lineage_scores )
```
Note that one of the columns is called id and contains the cell names. 

We can use the lineage scores and SingleCellExperiments as input to COSICC_DA_lineage. 

```{r}
lineage_result <- COSICC_DA_lineage(sce_DA_lineage_case,sce_DA_lineage_control,lineage_scores)
```

The output looks as follows:

```{r}
head(lineage_result)
```

